---
title: "The difference in mass between subspecies of common chaffinch, *Fringilla coelebs*"
author: "Emma Rand"
engine: knitr
execute:
  echo: false
  include: true
  error: false
  message: false
  warning: false
crossref:
  custom:
    - kind: float
      key: suppfig
      reference-prefix: Figure S
      space-before-numbering: false
    - kind: float
      key: suppfile
      reference-prefix: File S
      space-before-numbering: false
      caption-location: top
bibliography: references.bib
---

```{r}
#| label: packages
#| include: false
library(tidyverse)
library(broom)
```

```{r}
options(digits = 2)
```

# Introduction

A number of [subspecies of the common chaffinch, *Fringilla
coelebs*](https://en.wikipedia.org/wiki/Common_chaffinch), have been
described based principally on the differences in the pattern and colour
of the adult male plumage [@suárez2009]. Two of groups of these
subspecies are the "coelebs group" (@fig-coelebs) that occurs in Europe
and Asia and the "canariensis group" (@fig-canariensis) that occurs on
the Canary Islands. Or in Supplementary @suppfig-subspecies2!

::: {#fig-subspecies layout-ncol="2"}
![*F. c.
coelebs*](images/512px-Chaffinch_(Fringilla_coelebs).jpg){#fig-coelebs}

![*F. c.
palmae*](images/512px-Fringilla_coelebs_palmae_-_Los_Tilos.jpg){#fig-canariensis}

Adult male *Fringilla coelebs* of the coelebs group on the left (Andreas
Trepte, CC BY-SA 2.5 <https://creativecommons.org/licenses/by-sa/2.5>,
via Wikimedia Commons) and of the canariensis group on the right (H.
Zell, CC BY-SA 3.0 <https://creativecommons.org/licenses/by-sa/3.0>, via
Wikimedia Commons).
:::

This study sought to determine whether the two subspecies differ in
mass. The raw data are in @suppfile-chaff.

# Methods

```{r}
#| label: data-import
#| include: false
file <- "data-raw/chaff.txt"
chaff <- read_table(file)
```

```{r}
#| label: data-summary
chaff_summary <- chaff |> 
  group_by(subspecies) |>  
  summarise(mean = mean(mass, na.rm = TRUE),
            sd = sd(mass, na.rm = TRUE),
            n = length(mass),
            se = sd / sqrt(n))
```

```{r}
#| label: extract-summary
# extract values for inline reporting

# sample sizes
n_coelebs <- chaff_summary$n[chaff_summary$subspecies == "coelebs"] 
n_canariensis <- chaff_summary$n[chaff_summary$subspecies == "canariensis"] 

# highest mean, se and the subspecies to which they belong
highest_mean <- max(chaff_summary$mean) 
highest_se <- chaff_summary$se[chaff_summary$mean == highest_mean] 
highest_subspecies <- chaff_summary$subspecies[chaff_summary$mean == highest_mean]

# lowest mean, se and the subspecies to which they belong
lowest_mean <- min(chaff_summary$mean) 
lowest_se <- chaff_summary$se[chaff_summary$mean == lowest_mean] 
lowest_subspecies <- chaff_summary$subspecies[chaff_summary$mean == lowest_mean]
```

We randomly sampled `r n_coelebs` *F. c. coelebs* males and
`r n_canariensis` *F. c. palmae* males and determined their mass with
spring scales. Analysis was carried out with R version 4.3.1 [@R-core]
and tidyverse [@tidyverse] packages.

```{r}
#| label: model
# create a two-sample model and tidy result
mod <- lm(data = chaff, mass ~ subspecies) 
results <-  tidy(mod)

```

```{r}
#| label: check-assumptions
#| include: false
plot(mod, which = 1)

ggplot(mapping = aes(x = mod$residuals)) +
  geom_histogram(bins = 10)

# shapiro.test(mod$residuals)
# Conclusion: examination consitent with residuals being ND
# 
# -   Variance is about the same for all values of x
# -   Looks roughly normal, symmetrical and normality test is not significant

```

# Results

`r str_to_sentence(highest_subspecies)` ($\bar{x} \pm s.e.$:
`r highest_mean` $\pm$ `r highest_se` g) have significantly higher mass
than `r lowest_subspecies` (`r lowest_mean` $\pm$ `r lowest_se` g)($t$ =
`r abs(results$statistic[2])`; $d.f.$ = `r mod$df.residual`; $p$ =
`r results$p.value[2]`). See @fig-chaff

::: {#fig-chaff}
```{r}

ggplot() +
  geom_point(data = chaff, aes(x = subspecies, y = mass),
             position = position_jitter(width = 0.1, height = 0),
             colour = "gray50") +
  geom_errorbar(data = chaff_summary, 
                aes(x = subspecies, ymin = mean - se, ymax = mean + se),
                width = 0.3) +
  geom_errorbar(data = chaff_summary, 
                aes(x = subspecies, ymin = mean, ymax = mean),
                width = 0.2) +
  scale_y_continuous(name = "Mass (g)", 
                     limits = c(0, 30), 
                     expand = c(0, 0)) +
  scale_x_discrete(name = "Subspecies", 
                   labels = c("Canariensis", "Coelebs")) +
  annotate("segment", x = 1, xend = 2, 
           y = 28, yend = 28,
           colour = "black") +
  annotate("text", x = 1.5,  y = 29, 
           label = expression(italic(p)~"= 0.012")) +
  theme_classic()
```

**Canariensis chaffinches are heavier than Coelebs chaffinches**. The
mean mass of 20 randomly sampled males from each subspecies was
determined. Error bars are $\pm$ 1 standard error. Canariensis
chaffinches were significantly heavier than Coelebs ($t$ =
`r abs(results$statistic[2])`; $d.f.$ = `r mod$df.residual`; $p$ =
`r results$p.value[2]`). Data analysis was conducted in R [@R-core] with
tidyverse packages [@tidyverse].
:::

There is another figure showing the distribution of the data in
@fig-chaff-dist presented here only to demonstrate the auto numbering of
figures. The same figure is in "Supplemental Figures". See
@suppfig-chaff-dist2

```{r}
#| label: fig-chaff-dist
#| fig-cap: "Unnecessary extra figure to demonstrate the autonumbering."
#| fig-width: 3.5
#| fig-height: 3.5
#| fig-align: left
ggplot(data = chaff, aes(x = mass, fill = subspecies)) +
  geom_density(alpha = 0.4) +
    scale_y_continuous(name = "Density",
                     limits = c(0, 0.2),
                     expand = c(0, 0)) +
  scale_x_continuous(name = "Mass (g)",
                     limits = c(10, 30),
                     expand = c(0, 0)) +
  scale_fill_viridis_d(name = "Subspecies",
                      labels = c("Canariensis", 
                                 "Coelebs")) +
  theme_classic()
                   
```

These results are gratuitously reproduced in @tbl-chaff just to include
a table.

```{r}
#| label: tbl-chaff
chaff_summary |> 
  mutate(subspecies = str_to_sentence(subspecies)) |> 
  knitr::kable(caption = "Summary of mass(g) for the two subspecies of chaffinch.",
               col.names = names(chaff_summary) |> str_to_sentence())

```

I have created a mutli-column figure with mixed content. Take a look at
@fig-mixedcontent. You can see two histograms in @fig-hist and
@fig-hist2. There is also a photo of a chaffinch in @fig-coelebs2.

::: {#fig-mixedcontent layout-ncol="2"}
```{r}
#| label: fig-hist
chaff |> 
  filter(subspecies == "canariensis") |>
  ggplot(aes(x = mass)) +
  geom_density() +
  theme_minimal()
```

```{r}
#| label: fig-hist2
chaff |> 
  filter(subspecies == "coelebs") |>
  ggplot(aes(x = mass)) +
  geom_density() +
  theme_minimal()

```

![*F. c.
coelebs*](images/512px-Chaffinch_(Fringilla_coelebs).jpg){#fig-coelebs2}

```{r}
#| label: fig-table

# notice that this is a table but I had to use the fig- prefix 
# to get it to be treated like a figure rather than the tbl- prefix

chaff_summary |> 
  mutate(subspecies = str_to_sentence(subspecies)) |> 
  knitr::kable(caption = "Summary of mass(g) for the two subspecies of chaffinch.",
               col.names = names(chaff_summary) |> str_to_sentence())

```

A figure with a some plots and a photo. There's even a table in there
although, IMO tables should not be in figures, but placed separately. I
had to use the fig- prefix in the label to get make it be treated like a
figure with caption below etc.
:::

# Discussion

Here we pick up points from the introduction.

### References

::: {#refs}
:::

# Supplemental

## Files

::: {#suppfile-chaff}
[chaff.txt](data-raw/chaff.txt)
:::


::: {#suppfile-pdfreport}
[example-pdf.pdf](example-pdf.pdf)
:::



## 

::: {#suppfig-chaff-dist2}
```{r}
ggplot(data = chaff, aes(x = mass, fill = subspecies)) +
  geom_density(alpha = 0.4) +
    scale_y_continuous(name = "Density",
                     limits = c(0, 0.2),
                     expand = c(0, 0)) +
  scale_x_continuous(name = "Mass (g)",
                     limits = c(10, 30),
                     expand = c(0, 0)) +
  scale_fill_viridis_d(name = "Subspecies",
                      labels = c("Canariensis", 
                                 "Coelebs")) +
  theme_classic()
               
```

Here's that same figure again, but this time it's in the supplemental
material.
:::

::: {#suppfig-subspecies2 layout-ncol="2"}
![*F. c.
coelebs*](images/512px-Chaffinch_(Fringilla_coelebs).jpg){#suppfig-coelebs2}

![*F. c.
palmae*](images/512px-Fringilla_coelebs_palmae_-_Los_Tilos.jpg){#suppfig-canariensis2}

Adult male *Fringilla coelebs* of the coelebs group on the left (Andreas
Trepte, CC BY-SA 2.5 <https://creativecommons.org/licenses/by-sa/2.5>,
via Wikimedia Commons) and of the canariensis group on the right (H.
Zell, CC BY-SA 3.0 <https://creativecommons.org/licenses/by-sa/3.0>, via
Wikimedia Commons).
:::
